Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-08-28 manuscript/supplement_images.tex
progress on revision

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
\usepackage{setspace}
  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
\centering{\Large Paper title goes here \par}

\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}

\textbf{\centering{\large Supplementary materials \par}} \newline
Supplementary images S1 - S4

%%%%%%%%%%%%%%%%%
%% Figure S1
%%%%%%%%%%%%%%%%%
\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
\end{center}

\newpage
%%%%%%%%%%%%%%%%%
%% Figure S2
%%%%%%%%%%%%%%%%%
\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for the respective data type.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
	\caption{Figure S2}
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S3
%%%%%%%%%%%%%%%%%

\newpage

\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values are calculated using held out test data.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
	\caption{Figure S3}
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S4
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S4}: Distribution of model parameters for the three species common to the Hubbard Brook, Harvard, and NPN datasets. The phenophase is budburst for all three species. Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_hubbard_harvard_comparison.png}
	\caption{Figure S4}
\end{center}

>>>>

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
    \linenumbers
%\usepackage{setspace}
%  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}

{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline

\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White} \newline

{\Large \textbf{Supplementary materials}} \newline
Methods describing processing of the 4 LTER datasets

The four LTER datasets each had different protocols for recording phenology observations. Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network. As in the USA-NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation. The years used for each datasets were all years available at the time of analysis.

\textbf{Harvard Forest} \newline
We used observations from 1990-2014. Observations here were recorded as relative percentage of flowering or budburst for individual plants. We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%

\textbf{H.J. Andrews Experimental Forest} \newline
We used observations from 2009-2015. We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an indiviudal was first marked as "Flowers open". Note that each species has slightly different ordinal codes to mark each of these events.

\textbf{Jornada Experimental Range} \newline
We used observations from 1992-2009. Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase. We set "Yes" observations for flowers to the first DOY where the flower phenophase was 10\% or greater.

\textbf{Hubbard Brook} \newline
We used observations from 1989-2015. Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline

0: winter conditions \newline
1: bud swelling \newline
2: small leaves or flowers \newline
3: leafs 1/2 of final length, leafs obscure half the sky as seen thru crown \newline
3.5: leaves 3/4 expanded, sky mostly obscured, crown not yet in summer condition \newline
4: fully expanded, canopy in summer conditions \newline

We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].

### 2018-08-28 manuscript/supplement_methods.tex
progress on revision

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
    \linenumbers
%\usepackage{setspace}
%  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
\begin{center}
{\huge Paper title goes here \par} \newline

\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White} \newline

{\Large \textbf{Supplementary materials}} \newline
Methods describing processing of the 4 LTER datasets
\end{center}

The four LTER datasets each had different protocols for recording phenology observations. Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network. As in the NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation.

\textbf{Harvard Forest} \newline
Observations here were recorded as relative percentage of flowering or budburst for individual plants. "Yes" observations budburst and flower where set to the DOY when the percentage of each tree had was greater than or equal to 10\%

\textbf{H.J. Andrews Experimental Forest} \newline
"yes" observations for budburst was the first DOY when an individual was marked as "bud break". "yes" observations for flowering was was when an indiviudal was first marked as "Flowers open". Note that each species has slightly different ordinal codes to make each of these events.

\textbf{Jornada Experimental Range} \newline
Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase. "Yes" observations for flowers was the first DOY where the flower phenophase was 10\% or greater.

\textbf{Hubbard Brook} \newline
Observations for this dataset represent for each species the average, among 3 individuals, of an ordinal description of phenophase. \newline

0: winter conditions \newline
1: bud swelling \newline
2: small leaves or flowers \newline
3: leafs 1/2 of final length, leafs obscure half the sky as seen thru crown \newline
3.5: leaves 3/4 expanded, sky mostly obscured, crown not yet in summer condition \newline
4: fully expanded, canopy in summer conditions \newline

"Yes" observations for budburst were the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].

>>>>

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
\usepackage{setspace}
  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}

\textit{Shawn D. Taylor, Joan M. Meiners, Kristina Riemer, Michael C. Orr, Ethan P. White}

\textbf{\large Supplementary materials}

Supplementary images S1 - S7 and Table S1

%%%%%%%%%%%%%%%%%
%% Figure S1
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S1}: RMSE for specific species and phenophases of the USA-NPN dataset. X marks the best performing models for each respective data type.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
	Figure S1
\end{center}

\newpage
%%%%%%%%%%%%%%%%%
%% Figure S2
%%%%%%%%%%%%%%%%%
\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for each respective data type.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
	Figure S2
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S3
%%%%%%%%%%%%%%%%%

\newpage

\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.

\newpage

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
	Figure S3
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S4
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S4}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and USA-NPN datasets. The phenophase is budburst for all three species. Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution. Note the heading for each sub figure.

\newpage

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
	Figure S4
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S5
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S5}: As in Figure S4, but for the Alternating and Uniforc models.

\newpage

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
	Figure S5
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S6
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S6}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and USA-NPN derived models. The phenophase for the four species is budburst. These 4 species are representative of the analysis, and for the remaining comparisons the reader is pointed to the script 'analysis/plot\_select\_species\_parameters.R' in the code repository to generate additional figures.

\newpage

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
	Figure S6
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S7
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S7}: As in Figure S6, but for the Alternating and Uniforc models.

\newpage

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
	Figure S7
\end{center}
%%%%%%%%%%%%%%%%%
%% Table S1
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Table S1}: Species used in the analysis along with the sample size from each dataset. The numbers indicate the sample size of the training data and the size of the testing data in parenthesis.

\newpage

\begin{table}
\tiny
\begin{tabular}{l|r|l|l|l|l|l|l}
\hline
species & phenophase & phenophase\_type & harvard & hjandrews & hubbard & jornada & npn\\
\hline
acer circinatum & 371 & Budburst & - & 266 (66) & - & - & 39 (10)\\
\hline
acer circinatum & 501 & Flowers & - & 116 (29) & - & - & 33 (8)\\
\hline
acer pensylvanicum & 371 & Budburst & 80 (20) & - & - & - & 34 (9)\\
\hline
acer rubrum & 371 & Budburst & 100 (25) & - & - & - & 957 (239)\\
\hline
acer rubrum & 501 & Flowers & 96 (24) & - & - & - & 668 (167)\\
\hline
acer saccharum & 371 & Budburst & 60 (15) & - & 164 (41) & - & 365 (91)\\
\hline
betula alleghaniensis & 371 & Budburst & 60 (15) & - & 178 (44) & - & 133 (33)\\
\hline
betula alleghaniensis & 501 & Flowers & 26 (7) & - & - & - & 64 (16)\\
\hline
betula lenta & 371 & Budburst & 58 (14) & - & - & - & 96 (24)\\
\hline
betula papyrifera & 371 & Budburst & 76 (19) & - & - & - & 96 (26)\\
\hline
betula papyrifera & 501 & Flowers & 18 (5) & - & - & - & 36 (9)\\
\hline
fagus grandifolia & 371 & Budburst & 76 (19) & - & 177 (44) & - & 259 (65)\\
\hline
fraxinus americana & 371 & Budburst & 84 (21) & - & - & - & 90 (23)\\
\hline
fraxinus americana & 501 & Flowers & 22 (6) & - & - & - & 52 (13)\\
\hline
ilex verticillata & 371 & Budburst & 35 (9) & - & - & - & 26 (6)\\
\hline
larrea tridentata & 501 & Flowers & - & - & - & 27 (7) & 118 (30)\\
\hline
nyssa sylvatica & 371 & Budburst & 27 (7) & - & - & - & 63 (16)\\
\hline
pinus strobus & 496 & Budburst & 38 (10) & - & - & - & 77 (19)\\
\hline
populus tremuloides & 371 & Budburst & 38 (10) & - & - & - & 208 (51)\\
\hline
populus tremuloides & 501 & Flowers & 17 (4) & - & - & - & 79 (22)\\
\hline
prosopis glandulosa & 501 & Flowers & - & - & - & 49 (12) & 78 (20)\\
\hline
prunus serotina & 371 & Budburst & 58 (14) & - & - & - & 228 (57)\\
\hline
pseudotsuga menziesii & 480 & Budburst & - & 182 (46) & - & - & 38 (10)\\
\hline
quercus alba & 371 & Budburst & 62 (15) & - & - & - & 174 (43)\\
\hline
quercus rubra & 371 & Budburst & 80 (20) & - & - & - & 242 (60)\\
\hline
quercus rubra & 501 & Flowers & 56 (14) & - & - & - & 127 (32)\\
\hline
quercus velutina & 371 & Budburst & 77 (19) & - & - & - & 72 (18)\\
\hline
rhododendron macrophyllum & 371 & Budburst & - & 84 (21) & - & - & 48 (12)\\
\hline
rhododendron macrophyllum & 501 & Flowers & - & 27 (7) & - & - & 50 (12)\\
\hline
trillium ovatum & 488 & Budburst & - & 222 (55) & - & - & 68 (17)\\
\hline
trillium ovatum & 501 & Flowers & - & 169 (42) & - & - & 60 (15)\\
\hline
vaccinium corymbosum & 371 & Budburst & 38 (10) & - & - & - & 60 (15)\\
\hline
vaccinium corymbosum & 501 & Flowers & 38 (10) & - & - & - & 65 (16)\\
\hline
vaccinium parvifolium & 371 & Budburst & - & 149 (37) & - & - & 25 (6)\\
\hline
vaccinium parvifolium & 501 & Flowers & - & 162 (41) & - & - & 25 (6)\\
\hline
\end{tabular}
\newline
Table S1
\end{table}
